---
title: "PFOS semi-quantification report"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("C:/Users/karpa/OneDrive - Kruvelab/Helen_phd/projects_measurements/Lara_PFAS")
library(caTools)
library(tidyverse)
library(caret)
library(plotly)

lara_concentrations_pred <- readRDS("lara_PFAS_data.RData")


```

### Calibration graph

Predicted ionization efficiencies vs experimental response factors of calibration compounds
(regression line and 10-times error lines)

```{r, echo = FALSE, warning=FALSE, message=FALSE}
ggplotly(lara_concentrations_pred$plot_predictedIE_slope)

```



+ In the file of calibration compounds and their SMILES: 43 compounds (in analysis file found 42)  
Compound **"PFPeDA"** present in SMILES file but not found in analysis file

+ Changed the name of following compound that did not match in SMILES and analysis file:  
 **"9Cl-PF3ONS (F-35B)" --> "9Cl-PF3ONS"**

+ took out  **"MeFOSE"** and **"EtFOSE"** from the calibration compounds as they did not have retention time values

+ Compounds **"PFNS"**, **"PFHpS"** and **"PFPeS"** (up left on the plot) seemed to have lower areas compared to areas of other calibration compounds. Are they known to have low ionization efficiencies? I left them out from the regression line calculation right now.



### Calibration compounds - real vs predicted concentrations

```{r, echo = FALSE, warning=FALSE, message=FALSE}
ggplotly(lara_concentrations_pred$plot_predicted_theoretical_conc)

```

(both graphs are interactive - hover over points or double-click a compound in the legend to see respective graph for the compound)











